---
title: "IMC data analysis workflow"
author: "Tess Brodie"
date: "University of Bern, DBMR, Visceral and Transplantation Surgery lab"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Quality control cell mask

```{r setup, include=FALSE}
#Setup knitr, incuding path
knitr::opts_chunk$set(echo = TRUE)

#Important - this path is relative to your dataset, set working directory
knitr::opts_knit$set(root.dir = "C:/Users/IMC Platform laptop/Users2/Tess/SteinbockCellSegmentation/220707/220707_workflowr")

#Set cache
knitr::opts_chunk$set(autodep = TRUE, cache = TRUE, cache.lazy = FALSE)

#To make figures
knitr::opts_chunk$set(dev = "png")
```

```{css, echo=FALSE}
.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
  background-color: inherit;
}
```

### Load packages 

```{r load-libs, message=FALSE}

library(imcRtools)
library(cytomapper)
library(openxlsx)
library(stringr)
library(dittoSeq)
library(RColorBrewer)
library(cytomapper)
library(viridis)
library(tidyverse)
library(ggrepel)
library(EBImage)
library(scuttle)
library(mclust)
library(scater)
library(patchwork)
library(batchelor)
library(harmony)
library(Seurat)
library(SeuratObject)
library(cowplot)
```

### Load data

```{r load data}
spe <- readRDS(file.path("output", "spe0.rds"))
```

### Read in images and create metadata for images object

```{r reading in images, class.output="scroll-100"}
#Load all of the .tiff images from the img folder
images <- loadImages("./data/img")

#Load all of the masks from the masks folder
masks <- loadImages("./data/masks", as.is = TRUE)

#Assign the channel names of the "images" object based on the SPE structure
channelNames(images) <- rownames(spe)

#Visual control
images
all.equal(names(images), names(masks))

#Create metadata for the "images" object

meta <- read.xlsx("./data/sample_metadata.xlsx")

sample_id <-  c("1", "2", "3", "4", "5", "6")
patient_id <- str_extract_all(names(images), ".*[1-6]", simplify = TRUE)
condition <- meta$condition[match(sample_id, meta$Sample.ID)] 

mcols(images) <- mcols(masks) <- DataFrame(sample_id = c("1", "2", "3", "4", "5", "6"),
                                           patient_id = c("1", "2", "3", "4", "5", "6"),
                                           condition = condition)
mcols(images)
```

### Save rds

```{r read in data}
saveRDS(images, "data/images.rds")
saveRDS(masks, "data/masks.rds")
```

### Segmentation quality control

```{r segmentation QC, class.output="scroll-100"}
set.seed(123)
img_ids <- sample(seq_len(length(images)), 6)

# Normalize and clip images
cur_images <- images[img_ids]
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, inputRange = c(0,0.2))


plotPixels(cur_images,
           mask = masks[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("CD206", "E-cadherin", "DNA1", "DNA2", "HistoneH3"),
           colour = list("CD206" = c("black", "green"),
                         "E-cadherin" = c("black", "red"),
                         "DNA1" = c("black", "blue"), "DNA2"= c("black", "blue"),
                         "HistoneH3" = c("black", "blue")),
           bcg = list("DNA1" = c(0, 3, 1),
                      "DNA2" = c(0, 3, 1),
                      "HistoneH3" = c(0, 3, 1),
                      "E-cadherin" = c(0, 10, 1),
                      "CD206" = c(0, 10 ,1)),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))

```

### Visualize single-cell expression in a heatmap

```{r single cell expression heatmap}

cur_cells <- sample(seq_len(ncol(spe)), 2000)

dittoHeatmap(spe[,cur_cells], genes = rownames(spe),
             assay = "exprs", cluster_cols = TRUE, scale = "none",
             heatmap.colors = viridis(100), annot.by = "condition",
             annotation_colors = list(indication = metadata(spe)$color_vectors$indication))
```

### Image level QC: signal to noise ratio (SNR) for each marker

```{r heatmap qc, class.output="scroll-100"}
cur_snr <- lapply(images, function(img){
    mat <- apply(img, 3, function(ch){
        # Otsu threshold
        thres <- otsu(ch, range = c(min(ch), max(ch)))
        # Signal-to-noise ratio
        snr <- mean(ch[ch > thres]) / mean(ch[ch <= thres])
        # Signal intensity
        ps <- mean(ch[ch > thres])
        
        return(c(snr = snr, ps = ps))
    })
    t(mat) %>% as.data.frame() %>% 
        mutate(marker = colnames(mat)) %>% 
        pivot_longer(cols = c(snr, ps))
})

cur_snr <- do.call(rbind, cur_snr)

#Subsetting the SNR plot values to only contain channels of interest

#cur_snr <- cur_snr[cur_snr$marker %in% 
                     #(rownames(spe)[rowData(spe)$use_channel], "DNA1", "DNA2", "HistoneH3"), ]

cur_snr %>% 
    group_by(marker, name) %>%
    summarize(mean = mean(value),
              ci = qnorm(0.975)*sd(value)/sqrt(n())) %>%
    pivot_wider(names_from = name, values_from = c(mean, ci)) %>%
    ggplot() +
#    geom_errorbar(aes(y = log2(mean_snr), xmin = log2(mean_ps - ci_ps), 
#                      xmax = log2(mean_ps + ci_ps))) +
#    geom_errorbar(aes(x = log2(mean_ps), ymin = log2(mean_snr - ci_snr), 
#                      ymax = log2(mean_snr + ci_snr))) +
    geom_point(aes(log2(mean_ps), log2(mean_snr))) +
    geom_label_repel(aes(log2(mean_ps), log2(mean_snr), label = marker)) +
    theme_minimal(base_size = 15) + ylab("Signal-to-noise ratio") +
    xlab("Signal intensity")
```

### Image area covered by cells 

```{r cell coverage}
colData(spe) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    summarize(cell_area = sum(area),
           no_pixels = mean(width_px) * mean(height_px)) %>%
    mutate(covered_area = cell_area / no_pixels) %>%
    ggplot() +
        geom_point(aes(sample_id, covered_area)) + 
        theme_minimal(base_size = 15) +
        ylim(c(0, 1)) + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
        ylab("% covered area") + xlab("")
```


### Display single images with low cell counts to analyze

```{r display images, class.output="scroll-100"}
# Normalize and clip images
cur_images <- images[c("220525_segmentationkit_testonNASHmice_005", "220525_segmentationkit_testonNASHmice_006")]
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, inputRange = c(0, 0.2))

plotPixels(cur_images,
           mask = masks[c("220525_segmentationkit_testonNASHmice_005", "220525_segmentationkit_testonNASHmice_006")],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("CD206", "HNFa", "CD3e", "E-cadherin", "DNA1"),
           colour = list("CD206" = c("black", "yellow"),
                         "HNFa" = c("black", "red"),
                         "CD3e" = c("black", "green"),
                         "E-cadherin" = c("black", "cyan"),
                         "DNA1" = c("black", "blue")),
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
```

### Mean marker expression per image to identify images with outlying cell type compositions

```{r mean marker expression across samples, fig.width= 8, fig.height=5}
library(scuttle)

image_mean <- aggregateAcrossCells(spe, 
                                   ids = spe$sample_id, 
                                   statistics="mean",
                                   use.assay.type = "counts")

assay(image_mean, "exprs") <- asinh(counts(image_mean))

dittoHeatmap(image_mean, genes = rownames(spe),
             assay = "exprs", cluster_cols = TRUE, scale = "none",
             heatmap.colors = viridis(100), 
             annot.by = c("condition", "patient_id", "ROI"),
             annotation_colors = list(indication = metadata(spe)$color_vectors$indication,
                                      patient_id = metadata(spe)$color_vectors$patient_id,
                                      ROI = metadata(spe)$color_vectors$ROI),
             show_colnames = TRUE)
```

### Cell level QC: SNR for each marker

```{r SNR cell level, class.output="scroll-100"}
library(mclust)

set.seed(123)
mat <- apply(assay(spe, "exprs"), 1, function(x){
    cur_model <- Mclust(x, G = 2)
    mean1 <- mean(x[cur_model$classification == 1])
    mean2 <- mean(x[cur_model$classification == 2])
    
    signal <- ifelse(mean1 > mean2, mean1, mean2)
    noise <- ifelse(mean1 > mean2, mean2, mean1)
    
    return(c(snr = signal/noise, ps = signal))
})
    
cur_snr <- t(mat) %>% as.data.frame() %>% 
        mutate(marker = colnames(mat))

#Subsetting the SNR plot values to only contain channels of interest
#cur_snr <- cur_snr[cur_snr$marker %in% 
                     #c(rownames(spe)[rowData(spe)$use_channel], "DNA1", "DNA2", "Histon H3"), ]

cur_snr %>% ggplot() +
    geom_point(aes(log2(ps), log2(snr))) +
    geom_label_repel(aes(log2(ps), log2(snr), label = marker)) +
    theme_minimal(base_size = 15) + ylab("Signal-to-noise ratio") +
    xlab("Signal intensity")
```

### Distributions of cell size across the individual images

```{r cell size, class.output="scroll-100"}
colData(spe) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    ggplot() +
        geom_boxplot(aes(sample_id, area)) +
        theme_minimal(base_size = 15) + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
        ylab("Cell area") + xlab("")

#table with average cell size and areas:
summary(spe$area)

# remove very small cells from the dataset:
#sum(spe$area > 4) 
# remove them from the dataset
#spe <- spe[,spe$area >= 5]

```

### Percentage of image area covered by cells

```{r cells per mm2 }
colData(spe) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    summarize(cell_count = n(),
           no_pixels = mean(width_px) * mean(height_px)) %>%
    mutate(cells_per_mm2 = cell_count/(no_pixels/1000000)) %>%
    ggplot() +
        geom_point(aes(sample_id, cells_per_mm2)) + 
        theme_minimal(base_size = 15)  + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
        ylab("Cells per mm2") + xlab("")
```

### Differences in staining patterns

```{r staining QC, class.output="scroll-100", fig.width= 10, fig.height=15 }
#pdf("allchannels_multidimplot.pdf", width = 15, height = 20)
multi_dittoPlot(spe, vars = rownames(spe),
               group.by = "patient_id", plots = c("ridgeplot"), 
               assay = "exprs", 
               color.panel = metadata(spe)$color_vectors$patient_id)
#dev.off()
```

### UMAP and TSNE plots

```{r UMAP and TSNE}
spe <- runUMAP(spe, exprs_values = "exprs") 
spe <- runTSNE(spe, exprs_values = "exprs") 

# store data
reducedDims(spe)

# inspect data in UMAP
head(reducedDim(spe, "UMAP"))
# inspect data in TSNE
head(reducedDim(spe, "TSNE"))
```

### Assessment of potential “batch effects”

```{r compare UMAP and TSNE}
# visualize patient id 
p1 <- dittoDimPlot(spe, var = "patient_id", reduction.use = "UMAP", size = 0.2) + 
    scale_color_manual(values = metadata(spe)$color_vectors$patient_id) +
    ggtitle("Patient ID on UMAP")
p2 <- dittoDimPlot(spe, var = "patient_id", reduction.use = "TSNE", size = 0.2) + 
    scale_color_manual(values = metadata(spe)$color_vectors$patient_id) +
    ggtitle("Patient ID on TSNE")

# visualize region of interest id
p3 <- dittoDimPlot(spe, var = "ROI", reduction.use = "UMAP", size = 0.2) + 
    scale_color_manual(values = metadata(spe)$color_vectors$ROI) +
    ggtitle("ROI ID on UMAP")
p4 <- dittoDimPlot(spe, var = "ROI", reduction.use = "TSNE", size = 0.2) + 
    scale_color_manual(values = metadata(spe)$color_vectors$ROI) +
    ggtitle("ROI ID on TSNE")


(p1 + p3) / (p2 + p4)

```

### Save data

```{r save spe1}
saveRDS(spe, file.path("output", "spe1.rds"))
```
