---
title: "IMC data analysis workflow"
author: "Tess Brodie"
date: "University of Bern, DBMR, Visceral and Transplantation Surgery lab"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Dimensionality reduction and cell classification

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Important - this path is relative to your dataset, set working directory
knitr::opts_knit$set(root.dir = "C:/Users/IMC Platform laptop/Users2/Tess/SteinbockCellSegmentation/220707/220707_workflowr")

#Set cache
knitr::opts_chunk$set(autodep = TRUE, cache = TRUE, cache.lazy = FALSE)

#To make figures
knitr::opts_chunk$set(dev = "png")
```

```{css, echo=FALSE}
.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
  background-color: inherit;
}
```

### Load packages 

```{r load-libs, message=FALSE}
library(SpatialExperiment)
library(Rphenograph)
library(igraph)
library(dittoSeq)
library(viridis)
library(bluster)
library(BiocParallel)
library(ggplot2)
library(scran)
library(CATALYST)
library(kohonen)
library(ConsensusClusterPlus)
library(patchwork)
library(pheatmap)
library(gridExtra)
library(cytomapper)
library(Rphenoannoy)
library(SingleCellExperiment)
library(caret)
library(ggridges)
```

### Load data

```{r load data}
spe <- readRDS(file.path("output", "spe2.rds"))

# Sample cells
cur_cells <- sample(seq_len(ncol(spe)), 2000)
```

### Dimensionality reduction: Rphenoannoy

```{r Rphenoannoy}
mat <- t(assay(spe, "exprs"))

out <- Rphenoannoy(mat, k = 10)

clusters <- factor(membership(out[[2]]))

spe$pg_clusters <- clusters

dittoDimPlot(spe, var = "pg_clusters", 
             reduction.use = "UMAP", size = 0.2,
             do.label = TRUE) +
    ggtitle("Phenoannoy clusters expression on UMAP")
```

# Heatmap with phenoannoy clusters:

```{r create heatmap with phenoannoy data}
dittoHeatmap(spe[, cur_cells], 
             genes = rownames(spe),
             assay = "exprs", scale = "none",
             heatmap.colors = viridis(100), 
             annot.by = c("pg_clusters", "patient_id"),
             annot.colors = c(dittoColors(1)[1:length(unique(spe$pg_clusters))],
                              metadata(spe)$color_vectors$patient_id))
```

### Create a phenoannoy with FastMNN cell integration data 

```{r integrated cell phenoannoy}
set.seed(322)

mat <- reducedDim(spe, "fastMNN")

out <- Rphenoannoy(mat, k = 10)

clusters <- factor(membership(out[[2]]))

spe$pg_clusters_corrected <- clusters

dittoDimPlot(spe, var = "pg_clusters_corrected", 
             reduction.use = "UMAP_mnnCorrected", size = 0.2,
             do.label = TRUE) +
    ggtitle("Phenoannoy clusters expression on UMAP, integrated cells")
```

### Heatmap with FastMNN cell integrated phenoannoy

```{r create heatmap with integrated phenoannoy}
dittoHeatmap(spe[,cur_cells], 
             genes = rownames(spe),
             assay = "exprs", scale = "none",
             heatmap.colors = viridis(100), 
             annot.by = c("pg_clusters_corrected", "patient_id"),
             annot.colors = c(dittoColors(1)[1:length(unique(spe$pg_clusters_corrected))],
                              metadata(spe)$color_vectors$patient_id))
```

### Shared nearest neighbour graph: This takes a while to run

```{r shared nearest neigbhours graph, eval=FALSE, include=FALSE}
sam_cells <- sample(seq_len(ncol(spe)), 15000)

mat <- t(assay(spe, "exprs"))

combinations <- clusterSweep(mat[sam_cells,], BLUSPARAM=SNNGraphParam(),
                             k=c(10L, 20L), 
                             type = c("rank", "jaccard"), 
                             cluster.fun=c("walktrap", "louvain"),
                             BPPARAM = MulticoreParam(RNGseed = 123))

sil <- vapply(as.list(combinations$clusters), 
              function(x) mean(approxSilhouette(mat[sam_cells,], x)$width), 
              0)

ggplot(data.frame(method = names(sil),
                  sil = sil)) +
    geom_point(aes(method, sil)) +
    theme_classic(base_size = 15) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("Cluster parameter combination") +
    ylab("Average silhouette width")
```

### Best value for average silhouette width is louvain rank cluster k=20

### Phenograph with louvain community detection and rank-based weighting of edges (k=20)

```{r louvain phenograph}
clusters <- clusterCells(spe, 
                         assay.type = "exprs", 
                         BLUSPARAM = NNGraphParam(k = 20, 
                                                  cluster.fun = "louvain",
                                                  type = "rank",
                                                  BPPARAM = bpparam()))

spe$nn_clusters <- clusters

dittoDimPlot(spe, var = "nn_clusters", 
             reduction.use = "UMAP", size = 0.2,
             do.label = TRUE) +
    ggtitle("SNN clusters expression on UMAP")
```

### Heatmap louvain - rank - k=20

```{r heatmap louvain}
dittoHeatmap(spe[,cur_cells], 
             genes = rownames(spe),
             assay = "exprs", scale = "none",
             heatmap.colors = viridis(100), 
             annot.by = c("nn_clusters", "patient_id"),
             annot.colors = c(dittoColors(1)[1:length(unique(spe$nn_clusters))],
                              metadata(spe)$color_vectors$patient_id))
```

### Phenograph louvain with integrated cells

```{r louvain integrated phenograph}
clusters <- clusterCells(spe, 
                         use.dimred = "fastMNN", 
                         BLUSPARAM = NNGraphParam(k = 20, 
                                                  cluster.fun = "louvain",
                                                  type = "rank",
                                                  BPPARAM = bpparam()))

spe$nn_clusters_corrected <- clusters

dittoDimPlot(spe, var = "nn_clusters_corrected", 
             reduction.use = "UMAP_mnnCorrected", size = 0.2,
             do.label = TRUE) +
    ggtitle("SNN clusters expression on UMAP, integrated cells")
```

### Heatmap louvain integrated phenograph

```{r heatmap louvain integrated}
dittoHeatmap(spe[,cur_cells], 
             genes = rownames(spe),
             assay = "exprs", scale = "none",
             heatmap.colors = viridis(100), 
             annot.by = c("nn_clusters_corrected", "patient_id"),
             annot.colors = c(dittoColors(1)[1:length(unique(spe$nn_clusters_corrected))],
                              metadata(spe)$color_vectors$patient_id))
```

### flowSOM/Catalyst phenographs

```{r CATALYST phenograph}
# Run FlowSOM and ConsensusClusterPlus clustering
spe <- CATALYST::cluster(spe, 
               features = rownames(spe),
               maxK = 30)

# Assess cluster stability
delta_area(spe)
```

### Cell clusters with meta12

```{r metaclustering}
spe$som_clusters <- cluster_ids(spe, "meta12")

dittoDimPlot(spe, var = "som_clusters", 
             reduction.use = "UMAP", size = 0.2,
             do.label = TRUE) +
    ggtitle("SOM clusters expression on UMAP")
```

### Heatmap

```{r heatmap metaclusters}
dittoHeatmap(spe[,cur_cells], 
             genes = rownames(spe),
             assay = "exprs", scale = "none",
             heatmap.colors = viridis(100), 
             annot.by = c("som_clusters", "patient_id"),
             annot.colors = c(dittoColors(1)[1:length(unique(spe$som_clusters))],
                              metadata(spe)$color_vectors$patient_id))
```

### Clustering with bluster package

```{r bluster clustering}
# Select integrated cells
mat <- reducedDim(spe, "fastMNN")

# Perform SOM clustering
som.out <- clusterRows(mat, SomParam(100), full = TRUE)

# Cluster the 100 SOM codes into larger clusters
ccp <- ConsensusClusterPlus(t(som.out$objects$som$codes[[1]]),
                            maxK = 30,
                            reps = 100, 
                            distance = "euclidean", 
                            seed = 220410, 
                            plot = NULL)

CATALYST:::.plot_delta_area(ccp)
```

### Link ConsensusClusterPlus clusters with SOM codes and save in object

```{r bluster corrected}
som.cluster <- ccp[[16]][["consensusClass"]][som.out$clusters]
spe$som_clusters_corrected <- as.factor(som.cluster)

dittoDimPlot(spe, var = "som_clusters_corrected", 
             reduction.use = "UMAP_mnnCorrected", size = 0.2,
             do.label = TRUE) +
    ggtitle("Phenograph clusters expression on UMAP, integrated cells")
```

### Heatmap with SOM

```{r heatmap bluster corrected}
dittoHeatmap(spe[,cur_cells], 
             genes = rownames(spe),
             assay = "exprs", scale = "none",
             heatmap.colors = viridis(100), 
             annot.by = c("som_clusters_corrected", "patient_id"),
             annot.colors = c(dittoColors(1)[1:length(unique(spe$som_clusters_corrected))],
                              metadata(spe)$color_vectors$patient_id))
```

### Comparison between clustering approaches

```{r approach comparison}
tab1 <- table(paste("Rphenograph", spe$pg_clusters), 
              paste("SNN", spe$nn_clusters))
tab2 <- table(paste("Rphenograph", spe$pg_clusters), 
              paste("SOM", spe$som_clusters))
tab3 <- table(paste("SNN", spe$nn_clusters), 
              paste("SOM", spe$som_clusters))

ph1 <- pheatmap(log10(tab1 + 10), color = viridis(100))[[4]]

ph2 <- pheatmap(log10(tab2 + 10), color = viridis(100))[[4]]

ph3 <- pheatmap(log10(tab3 + 10), color = viridis(100))[[4]]

grid.arrange(ph1, ph2, ph3, nrow = 1, ncol = 3)
```

### Comparison between clustering approaches: integrated cells

```{r approach comparison cell integration}
tab1 <- table(paste("Rphenograph", spe$pg_clusters_corrected), 
              paste("SNN", spe$nn_clusters_corrected))
tab2 <- table(paste("Rphenograph", spe$pg_clusters_corrected), 
              paste("SOM", spe$som_clusters_corrected))
tab3 <- table(paste("SNN", spe$nn_clusters_corrected), 
              paste("SOM", spe$som_clusters_corrected))

ph1 <- pheatmap(log10(tab1 + 10), color = viridis(100))[[4]]

ph2 <- pheatmap(log10(tab2 + 10), color = viridis(100))[[4]]

ph3 <- pheatmap(log10(tab3 + 10), color = viridis(100))[[4]]

grid.arrange(ph1, ph2, ph3, nrow = 1, ncol = 3)
```

### Save data

```{r save spe3}
saveRDS(spe, file.path("output", "spe3.rds"))
```


